Drought, vegetation productivity, and land cover change in Chile

### 2024-03-12 article/paper_drought_Chile2023.tex
update paper qm intro

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Drought indices of water demand and supply and their impact on LULCC in Chile from ERA5-Land and MODIS},
  pdfauthor={Francisco Zambrano},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Drought indices of water demand and supply and their impact on
LULCC in Chile from ERA5-Land and MODIS}
\author[1]{Francisco Zambrano%
\corref{cor1}%
\fnref{fn1}}
 \ead{francisco.zambrano@umayor.cl}

\affiliation[1]{organization={Universidad Mayor, Hémera Centro de
Observación de la Tierra, Facultad de Ciencias, Escuela de Ingeniería en
Medio Ambiente y Sustentabilidad},city={Santiago,
Chile},postcode={7500994},postcodesep={}}

@@

\begin{abstract}
Central Chile has been the focus of research studies due to the
persistent decrease in water supply, which is impacting the hydrological
system and vegetation development. This persistent period of water
scarcity has been defined as a ``Mega Drought''. Our objective is to
examine the effects of drought on LULCC (land use land cover change)
over continental Chile using drought indices of water supply and demand,
soil moisture, and vegetation productivity. For the analysis,
continental Chile was divided into five zones according to a latitudinal
gradient: ``Norte Grande,'' ``Norte Chico,'' ``Centro,'' ``Sur,'' and
``Austral.'' The monthly ERA5-Land (ERA5L) variables for precipitation,
temperature, and soil moisture were used. From 2001 to 2022, we used the
land cover MODIS product MCD12Q1, and from 2000 to 2023, we used the
NDVI (Normalized Difference Vegetation Index) product MOD13A3 collection
6.1. We estimated atmospheric evaporative demand (AED) using the
Hargreaves-Samani equation with the ERA5L temperature. We used the
Standardized Precipitation Index (SPI), the Standardized Precipitation
Evapotranspiration Index (SPEI), the Evaporative Demand Drought Index
(EDDI), the Standardized Soil Moisture Index (SSI), and the Standardized
anomaly of cumulative NDVI (zcNDVI) as drought indicators. These indices
were calculated for time scales of 1, 3, 6, 12, 24, and 36 months,
except for zcNDVI, which was for 6 months. We analyze the trend for
LULCC, vegetation productivity, and drought indices. Also, we analyzed
the temporal correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to gain
insights into the impact of water supply and demand on vegetation
productivity. Our results showed that LULCC were highest in ``Centro,''
``Sur,'' and ``Austral,'' with 36\%, 31\%, and 34\%, respectively. The
EDDI shows that water demand has increased for all zones, with a major
increase in ``Norte Grande.'' The drought indices of water supply and
soil moisture evidence a decreasing trend, which decreases at longer
time scales, from ``Norte Grande'' to ``Sur.'' ``Austral'' is the only
zone that shows an increase in supply. Vegetation productivity measures
by zcNDVI present a negative trend in ``Norte Chico'' and ``Centro.''
Showing to be the zones most impacted by climatic conditions, the years
2020 and 2022 suffered the most extreme drought. On the other hand,
forests seem to be the most resistant to drought. The types that show to
be most affected by variation in climate conditions are shrublands,
savannas, and croplands. The drought indices that have the capability of
explaining to a major degree the variance in vegetation productivity are
SSI-12, followed by SPEI-24 and SPEI-12 in ``Norte Chico'' and
``Centro.'' The results indicate that ``Norte Chico'' and ``Zona
Central'' are the most sensitive regions to water supply deficits longer
than a year, potentially explained by a low capacity of water storage in
those zones that should be further investigated.
\end{abstract}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, interior hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

The sixth assessment report (AR6) of the IPCC \citep{IPCC2023} indicates
that human-induced greenhouse gas emissions have increased the frequency
and/or intensity of some weather and climate extremes, and the evidence
has been strengthened since AR5 \citep{IPCC2013}. There is a high degree
of confidence that rising temperatures will increase the land area where
droughts will occur more frequently and with greater severity
\citep{IPCCCH112021}. Furthermore, drought increases tree mortality and
triggers changes in land cover and, consequently, land use, thus
impacting ecosystems \citep{Crausbay2017}. Nevertheless, there is a lack
of understanding of how the alteration in water supply and demand is
affecting land cover transformations.

The primary cause of drought is precipitation, and temperature makes it
worse \citep{Luo2017}. Drought impacts soil moisture, hydrological
regimes, and vegetation productivity. Initially, drought was commonly
classified as meteorological, hydrological, and agricultural
\citep{Wilhite1985}. Lately, \citep{Loon2016} and
\citep{AghaKouchak2021} have given an updated definition of drought for
the Anthropocene, suggesting that it should be considered the feedback
of humans' decisions and activities that drives the anthropogenic
drought. Even though it has been argued that those definitions do not
fully address the ecological dimensions of drought, \citep{Crausbay2017}
proposed the ecological drought definition as ``an episodic deficit in
water availability that drives ecosystems beyond thresholds of
vulnerability, impacts ecosystem services, and triggers feedback in
natural and/or human systems.''. Moreover, many ecological studies have
misinterpreted how to characterize drought, for example, sometimes
considering ``dry'' conditions as ``drought'' \citep{Slette2019}. On the
other hand, the AR6 \citep{IPCC2023} predicts that many regions of the
world will experience more severe agricultural and ecological droughts
even if global warming stabilizes at 1.5°--2°C. Then, there is a
challenge in conducting drought research, especially to evaluate its
impact on ecosystems.

Chile has been facing a persistent rainfall deficit for more than a
decade \citep{Garreaud2017}, which has impacted vegetation development
\citep{Zambrano2023} and the hydrological system \citep{Boisier2018}.
Current drought conditions have affected crop productivity
\citep{Zambrano2016, Zambrano2018}, forest development
\citep{Miranda2020, Venegas2018}, forest fire occurrence
\citep{UrrutiaJalabert2018}, land cover change \citep{Fuentes2021},
water supply in watersheds \citep{AlvarezGarreton2021}, and have had
economic impacts \citep{Fernandez2023}. In 2019--2020, the drought
severity reached an extreme condition in Central Chile (30--34°S) not
seen for at least 40 years, and the evidence indicates that the impact
is transversal to the land cover classes of forest, grassland, and
cropland \citep{Zambrano2023}. The prolonged lack of precipitation in
Central Chile is producing changes in ecosystem dynamics that must be
studied.

For the spatiotemporal assessment of drought impact (i.e., by water
supply and demand) on land cover changes, we need climatic realiable
variables such as precipitation, temperature, soil moisture, land cover,
and vegetation status. For developing countries like Chile, the weather
networks present several disadvantages, such as gaps, a short history,
and low-quality data. Reanalysis data, as the ERA5-Land (ERA5L)
\citep{MunozSabater2021} provides hourly climatic information
(precipitation, temperature, and soil moisture) without gaps since 1950
with global extension. ERA5L has already been used for drought
assessment using the Standardized Precipitation-Evapotranspiration Index
(SPEI) \citep{Nouri2023} and for flash drought \citep{Wang2023} by
analyzing soil moisture and evapotranspiration. On the other hand,
satellite remote sensing \citep{West2019, AghaKouchak2015} is the
primary method to evaluate how drought impacts vegetation dynamics. The
Moderate-Resolution Imaging Spectroradiometer (MODIS) can be used to get
vegetation drought indices (VDI), which are often used as proxies for
productivity \citep{Paruelo2016, Schucknecht2017}. Besides, land use and
land cover (LULC) change can be driven by drought
\citep{Tran2019, Akinyemi2021}. To analyze these changes, multiple LULC
products exist \citep{Grekousis2015}. One of those that provides time
series since 2001 is the MCD12Q1 \citep{Friedl2019} from MODIS. The
variation in water supply and demand is finally reflected in the total
water storage (TWS). The Gravity Recovery and Climate Experiment
(GRACE), which allows analyzing changes in water availability at coarse
resolution, can retrieve the TWS \citep{Ahmed2014, Ma2017}. We can find
drought indices of supply (i.e., precipitation) and demand (i.e.,
temperature) using climatic reanalysis (ERA5L) and vegetation data
(MODIS). This lets us figure out how drought changes LULC. Further, the
TWS can be assessed with regard to the changes in water supply and
demand to gain insight into the impact on water storage.

To evaluate meteorological drought (i.e., water supply), the World
Meteorological Organization (WMO; \citep{WMO2012}) recommends the
Standardized Precipitation Index (SPI; \citep{Mckee1993}), a multiscalar
drought index that allows to monitor precipitation deficits from short-
to long-term. Following the same approach, \citep{Vicente-Serrano2010}
incorporates into the SPI the effect of temperature through the use of
potential evapotranspiration, thus proposing the SPEI (Standardized
Precipitation Evapotranspiration Index). Similarly, to evaluate solely
the evaporative demand driven by temperature, \citep{Hobbins2016} and
\citep{McEvoy2016} came up with the Evaporative Demand Drought Index
(EDDI). For vegetation, in a similar manner as the SPI, SPEI and EDDI;
\citep{Zambrano2018} proposed the zcNDVI, a standardized anomaly of the
cumulative Normalized Difference Vegetation Index (NDVI), which could be
acumulated over the growing season or any period (e.g., months),
resulting in a multiscalar drought index. For soil moisture, several
drought indices exist, such as the Soil Moisture Deficit Index (SDMI) a
normalized index \citep{Narasimhan2005} and the Soil Moisture
Agricultural Drought Index (SMADI) \citep{Souza2021} which is a
normalized index using vegetation, land surface temperature, and a
vegetation condition index (VCI, \citep{Kogan1995}). From TWS, we can
estimate the standardized terrestrial water storage index (STI)
\citep{Cui2021}, a standardized anomaly that follows the methodology of
the SPI, SPEI, EDDI, and zcNDVI. Thereby, we have drought indices for
water supply, demand, and storage, which can help to make a
comprehensive assessment of drought.

In this research, we aim to analyze the impact of drought on different
types of ecosystems (land cover classes) in continental Chile. Our
specific goals are: i) to analyze the trend on multi-scalar drought
indices for water demand and supply, soil moisture, and vegetation
productivity for 1981--2023/2001--2023; ii) to assess the LULC change
for 2001--2022 and how it relates to drought indices; iii) to evaluate
the relationship between zcNDVI, a measure of vegetation productivity,
and drought indices for water demand and supply and soil moisture; and
iv) to assess if the observed changes in the drought indices are linked
to TWS.

@@

Continetal Chile has a diverse climate condition from north to south and
east to west \citep{Aceituno2021} (Figure~\ref{fig-studyArea}), which
determines its great ecosystem diversity (Figure~\ref{fig-LCprop}). The
Andes Mountains are a main factor in latitudinal variation
\citep{Garreaud2009}. To describe the climate and ecosystem of Chile, we
use the Koppen-Geiger release by \citep{Beck2023} and the land cover
type persistance of 80\% of times for 2001--2022, from the IGBP
classification scheme \citep{Friedl2019} (see
Section~\ref{sec-methods_lulc}). ``Norte Grande'' and ``Norte Chico''
predominate in an arid desert climate with hot (Bwh) and cold (Bwk)
temperatures. At the south of ``Norte Chico,'' the climate changes to an
arid steppe with cold temperatures (Bsk). Mainly, the land is barren,
with a minor surface of vegetation types such as shrubland and
grassland. In the zones ``Centro'' and the north half of ``Sur,'' the
main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
There is a significant amount (50\%) of Chilean matorral (shrubland and
savanna, \citep{Fuentes2021}), then grassland (16\%), forest (8\%), and
croplands (5\%), in ``Centro.'' The south part of ``Sur'' and the north
part of ``Austral'' are dominated by an oceanic climate (Cfb). Those
zones are high in forest and grassland. The southern part of the country
has a tundra climate, and in Patagonia, it is a cold semi-arid area with
an extended surface of grassland, forest, and, to a lesser extent,
savanna.

@@

\caption{\label{fig-studyArea}(a) Chile with the Koppen-Geiger climate
classes and the five macrozones ``Norte Grande'', ``Norte Chico'',
``Centro'', ``Sur'', and ``Austral''. (b) Topography reference map. (c)
Land cover classes for 2022. (d) Persistent land cover classes
(\textgreater{} 80\%) for 2001-2022}

@@

For the downloading, processing, and analysis of the spatio-temporal
data, we used the open source software for statistical computing and
graphics, \texttt{R} \citep{R2023}. For downloading ERA5L, we used the
\texttt{\{ecmwfr\}} package \citep{Hufkens2019}. For processing raster
data, we used \texttt{\{terra\}} \citep{Hijmans2023} and
\texttt{\{stars\}} \citep{Pebesma2023}. For managing vectorial data, we
used \texttt{\{sf\}} \citep{Pebesma2018}. For the calculation of AED, we
used \texttt{\{SPEI\}} \citep{Bergueria2023}.

@@

To derive a proxy of vegetation productivity, we used the product
MOD13A3 collection 6.1 from MODIS \citep{Didan2015}. It provides
vegetation indices (NDVI and EVI) at 1km of spatial resolution and
monthly frequency. The MOD13A3.061 and MCD12Q1.061 were retrieved from
the online Data Pool, courtesy of the NASA EOSDIS Land Processes
Distributed Active Archive Center (LP DAAC), USGS Earth Resources
Observation and Science (EROS) Center, Sioux Falls, South Dakota,
https://lpdaa.usgs.gov/tools/data-pool/.

@@

To evaluate water demand, we chose the \(EDDI\)
\citep{Hobbins2016, McEvoy2016} index, which uses the \(AED\). For
supply, we used the index recommended by the World Meteorological
Organization (WMO) for monitoring drought, the SPI \citep{Mckee1993}. We
calculated the SPEI, which used a balance between \(P\) and \(AED\), in
this case, an auxiliary variable \(D = P-AED\) is used. In this study,
we used the \(SSI\) (standardized soil moisture index at 1 m)
\citep{Hao2013, AghaKouchak2014}, which uses soil moisture at 1m depth.
Finally, for the proxy of productivity, \(zcNDVI\), we used the NDVI.
Before using the NDVI, it was smoothed using a locally-weighted
polynomial regression, following the procedure described in
\citep{Zambrano2018} and \citep{Zambrano2016}.

@@

The time series of NDVI is affected by climatic conditions, vegetation
development, seasonality, and changes in vegetation type. In this study,
we want to analyze the variation in vegetation productivity in different
land cover types and how it is affected by water demand, water supply,
and soil moisture. In order to avoid changes due to a change in the land
cover type, that will wrongly impact NDVI. We will develop a persistence
mask for land cover for 2001--2023. Thereby, we reduce an important
source of variation on a regional scale.

>>>>

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Drought, vegetation productivity, and land cover change in Chile},
  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Fernández; Francisco Meza; Alejandro Venegas-González; Iongel Duran-Llacer; Nicolas Raab; Dylan Craven},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Drought, vegetation productivity, and land cover change in Chile}
\author[1]{Francisco Zambrano%
\corref{cor1}%
\fnref{fn1}}
 \ead{francisco.zambrano@umayor.cl} 
\author[2]{Anton Vrieling%
%
}
 \ead{a.vrieling@utwente.nl} 
\author[3]{Francisco Fernández%
%
}
 \ead{a.vrieling@utwente.nl} 
\author[]{Francisco Meza%
%
}

\author[]{Alejandro Venegas-González%
%
}

\author[]{Iongel Duran-Llacer%
%
}

\author[]{Nicolas Raab%
%
}

\author[]{Dylan Craven%
%
}

\affiliation[1]{organization={Universidad Mayor, Hémera Centro de
Observación de la Tierra, Facultad de Ciencias, Escuela de Ingeniería en
Medio Ambiente y Sustentabilidad},city={Santiago,
Chile},postcode={7500994},postcodesep={}}
\affiliation[2]{organization={University of Twente, Faculty of
Geo-Information Science and Earth Observation (ITC)},city={Enschede, The
Netherlands},postcodesep={}}
\affiliation[3]{organization={Universidad San
Sebastian},,postcodesep={}}

@@

\begin{abstract}
Central Chile has been the focus of research studies due to the
persistent decrease in water supply, which is impacting the hydrological
system and vegetation development. This persistent period of water
scarcity has been defined as a megadrought. Our objective is to assess
the impact of drought on LULCC (land use land cover change) over
continental Chile using drought indices of water supply and demand, soil
moisture, and their impact on vegetation productivity. The monthly
ERA5-Land (ERA5L) variables for precipitation, temperature, and soil
moisture were used. From 2001 to 2022, we used the land cover MODIS
product MCD12Q1, and from 2000 to 2023, we used the NDVI (Normalized
Difference Vegetation Index) product MOD13A3 collection 6.1. As drought
indices, we compute the standardized anomaly of cumulative NDVI
(zcNDVI), the Standardized Precipitation Evapotranspiration Index
(SPEI), the Evaporative Demand Drought Index (EDDI), and the
Standardized Soil Moisture Index (SSI). These indices were calculated
for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI,
which was for 6 months. We analyze the trend for LULCC, vegetation
productivity, and drought indices. Also, we analyzed the temporal
correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to gain insights
into the impact of water supply and demand on vegetation productivity.
Our results showed that LULCC was highest in ``Centro,'' ``Sur,'' and
``Austral,'' with 36\%, 31\%, and 34\% of change in the surface type,
respectively. The EDDI shows that water demand has increased for all
zones, with a major increase in ``Norte Grande.'' The drought indices of
water supply and soil moisture evidence a decreasing trend, which
decreases at longer time scales, from ``Norte Grande'' to ``Sur.''
``Austral'' is the only zone that shows an increase in supply.
Vegetation productivity measures by zcNDVI present a negative trend in
``Norte Chico'' and ``Centro.'' On the other hand, forests seem to be
the most resistant to drought. The types that show to be most affected
by variation in climate conditions are shrublands, savannas, and
croplands. The drought indices that have the capability of explaining to
a major degree the variance in vegetation productivity are SSI-12,
followed by SPEI-24 and SPEI-12 in ``Norte Chico'' and ``Centro.'' The
results indicate that ``Norte Chico'' and ``Zona Central'' are the most
sensitive regions to water supply deficits lasting longer than a year.
Our results can help develop a robust vegetation productivity
forecasting model for land cover classes in Chile.
\end{abstract}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, boxrule=0pt, frame hidden, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

Drought is often classified as meteorological when there is a decrease
in precipitation below the mean average of several years (more than 30
years), hydrological when these anomalies last for long periods (months
to years) and affect water systems, and agricultural when the deficit
impacts plant health anomalies and leads to decreased productivity
\citep{Wilhite1985}. However, it is important to note that drought is
also influenced by human activities, which were not considered in the
definitions. Thus, \citet{Loon2016} and \citet{AghaKouchak2021} have
given an updated definition of drought for the Anthropocene, suggesting
that it should be considered the feedback of humans' decisions and
activities that drives the anthropogenic drought. Simultaneously,
drought leads to heightened tree mortality and induces alterations in
land cover and land use, ultimately affecting ecosystems
\citep{Crausbay2017}. Even though many ecological studies have
misinterpreted how to characterize drought, for example, sometimes
considering ``dry'' conditions as ``drought'' \citep{Slette2019}. Then,
\citet{Crausbay2017} proposed the ecological drought definition as ``an
episodic deficit in water availability that drives ecosystems beyond
thresholds of vulnerability, impacts ecosystem services, and triggers
feedback in natural and/or human systems.'' In light of current global
warming, it is crucial to study the interaction between drought and
ecosystems in order to understand their feedback and impact on water
security. \citep{Bakker2012}

Human-induced greenhouse gas emissions have increased the frequency
and/or intensity of drought as a result of global warming, according to
the sixth assessment report (AR6) of the Intergovernmental Panel on
Climate Change (IPCC) \citep{IPCC2023}. The evidence supporting this
claim has been strengthened since AR5 \citep{IPCC2013}. Recent studies,
however, have produced contrasting findings, suggesting that drought has
not exhibited a significant trend over the past forty years.
\citep{Vicente-Serrano2022, Kogan2020}. \citet{Vicente-Serrano2022}
analyzed the meteorological drought trend on a global scale, finding
that only in a few regions has there been an increase in the severity of
drought. Moreover, they attribute the increase in droughts over the past
forty years solely to an increase in atmospheric evaporative demand
(AED), which in turn enhances vegetation water demand, with important
implications for agricultural and ecological droughts. Also, they state
that ``the increase in hydrological droughts has been primarily observed
in regions with high water demand and land cover change''. Similarly,
\citet{Kogan2020} analyzed the drought trend using vegetation health
methods, finding that for the globe, hemispheres, and main
grain-producing countries, drought has not expanded or intensified for
the last 38 years. Further, the \citet{IPCC2021} suggests that there is
a high degree of confidence that rising temperatures will increase the
extent, frequency, and severity of droughts. Also, AR6 \citep{IPCC2023}
predicts that many regions of the world will experience more severe
agricultural and ecological droughts even if global warming stabilizes
at 1.5°--2°C. To better evaluate the impact of drought trends on
ecosystems, assessments are needed that relate meteorological and soil
moisture variables to their effects on vegetation.

From 1960 to 2019, land use change has impacted around one-third of the
Earth's surface, which is four times more than previously thought
\citep{Winkler2021}. Multiple studies aim to analyze and forecast
changes in land cover globally \citep{Winkler2021, Song2018} and
regionally \citep{Chamling2020, Homer2020, Yang2021}. Some others seek
to analyze the impact of land cover change on climate conditions such as
temperature and precipitation \citep{Luyssaert2014, Pitman2012}. There
is less research on the interaction between drought and land cover
change \citep{Chen2022, Akinyemi2021, Peng2017}. \citet{Peng2017}
conducted a worldwide investigation utilizing net primary production to
examine the spatial and temporal variations in vegetation productivity
at global level. The study aimed to assess the influence of drought by
comparing the twelve-month Standardized Precipitation Evapotranspiration
Index (SPEI) and land cover change. According to their findings, drought
is responsible for 37\% of the decline in vegetation productivity, while
water availability accounts for 55\% of the variation. \citet{Chen2022}
studied the trend of vegetation greenness and productivity and its
relation to meteorological drought (SPEI of twelve months in December)
and soil moisture at the global level. The results showed lower
correlations (\textless0.2) for both variables. \citet{Akinyemi2021}
evaluates drought trends and land cover change using vegetation indices
in Botwsana in a semi-arid climate. These studies mostly looked at how
changes in land cover and vegetation productivity are related to a
single drought index (SPEI) over a single time period of 12 months. SPEI
takes into account the combined effect of precipitation and AED as a
water balance, but it does not allow us to know the contribution of each
variable on its own. Some things worth investigating in terms of land
cover change and vegetation productivity are: i) How do they respond to
short- to long-term meteorological droughts? ii) How do they behave in
humid and arid climatic zones regarding drought? And iii) What is the
role of soil moisture? Likewise, there is a lack of understanding of how
the alteration in water supply and demand is affecting land cover
transformations.

Chile's diverse climatic and ecosystem types
(\citet{Beck2023};\citet{Luebert2022}) make it an ideal natural
laboratory for studying climate and ecosystems. Additionally, the
country has experienced severe drought conditions that have had
significant effects on vegetation and water storage. Central Chile faced
a persistent precipitation deficit between 2010 and 2022, defined as a
megadrought \citep{Garreaud2017}, which has impacted the Chilean
ecosystem. This megadrought was defined by the Standardized
Precipitation Index (SPI) of twelve months in December having values
below one standard deviation. Some studies have addressed how this
drought affects single ecosystems in terms of forest development
\citep{Miranda2020, Venegas2018}, forest fire occurrence
\citep{UrrutiaJalabert2018}, and crop productivity
\citep{Zambrano2023, Zambrano2018, Zambrano2016}. We found one study
regarding land cover and drought in Chile. The study by
\citet{Fuentes2021} evaluates water scarcity and land cover change in
Chile between 29° and 39° of south latitude. \citet{Fuentes2021} used
the SPEI of one month for evaluating drought, which led to misleading
results. For example, they did not find a temporal trend in the SPEI but
found a decreasing trend in water availability and an increase trend on
AED, which in turn should have been capable of being captured with
longer time scales of the SPEI. The term ``megadrought'' in Chile is
used to describe a prolonged water shortage that lasts for several
years, resulting in a permanent deficit that impacts the hydrological
system \citep{Boisier2018}. Hence, it is imperative to assess temporal
scales that take into account the cumnulative effect within some years.
There is little knowledge about the relationship between drought and
ecosystem in Chile; thus, it is important to understand in more detail
how meteorological and soil moisture droughts influence ecosystem
dynamics to inform adaptation options.

A detailed spatiotemporal assessment of the interaction of drought for
short- to long-term and land cover change requires information on
vegetation as well as weather variables such as precipitation,
temperature, and soil moisture. Weather networks in Chile present some
disadvantages, such as spatio-temporal gaps, a short history, and
irregular quality, which make them difficult to represent the whole
extent of the country spatially. In order to do this, we use reanalysis
data from ERA5-Land \citep{MunozSabater2021} to create drought indices
that consider AED, precipitation, and soil moisture over a range of time
periods, from the short to the long term. Also, we use vegetation
spectral information and annual land cover change from the
Moderate-Resolution Imaging Spectroradiometer (MODIS). We expect to gain
insight regarding the temporal evolution of water demand, water supply,
and soil moisture, as well as the interaction with land cover change and
vegetation productivity. Here, we analyze the multi-dimensional impacts
of drought across ecosystems in continental Chile. More specifically, we
aim to assess: i) temporal changes in land-use cover and the direction
and magnitude of their relationships with drought indices for water
demand and supply, soil moisture, and vegetation productivity; ii)
short- to long-term temporal trends in multi-scalar drought indices; and
iii) the relationship between vegetation productivity and drought
indices for water demand and supply and soil moisture across Chilean
ecosystems.

@@

Continental Chile has a diverse climate conditions with strong gradients
from north to south and east to west \citep{Aceituno2021}
(Figure~\ref{fig-studyArea} a), which determines its great ecosystem
diversity (\citet{Luebert2022}) (Figure~\ref{fig-studyArea} c). The
Andes Mountains are a main factor in climate latitudinal variation
\citep{Garreaud2009}. In order to characterize the climate and ecosystem
of Chile, we utilize the Köppen-Geiger classification system developed
by \citet{Beck2023} and the land cover data derived from the MODIS
product for the period of 2001--2022, based on the International
Geosphere-Biosphere Programme (IGBP) classification scheme proposed by
\citet{Friedl2019}. ``Norte Grande'' and ``Norte Chico'' predominate in
an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At
the south of ``Norte Chico,'' the climate changes to an arid steppe with
cold temperatures (Bsk). In these two northern regions, the land is
mostly bare, with a minor surface of vegetation types such as shrubland
and grassland. In the zones ``Centro'' and the north half of ``Sur,''
the main climate is Mediterranean, with warm to hot summers (Csa and
Csb). Land cover in ``Centro'' comprises a significant amount of
shrubland and savanna (50\%), grassland (16\%), forest (8\%), and
croplands (5\%). An oceanic climate (Cfb) predominates in the south of
``Sur'' and the north of ``Austral.'' Those zones are high in forest and
grassland. The southern part of the country has a tundra climate, and in
``Austral'', it is a cold semi-arid area with an extended surface of
grassland, forest, and, to a lesser extent, savanna.

@@

\caption{\label{fig-studyArea}(a) Chile with the Koppen-Geiger climate
classes and the five macrozones ``Norte Grande'', ``Norte Chico'',
``Centro'', ``Sur'', and ``Austral''. (b) Topography reference map. (c)
land cover classes for 2022. (d) Persistent land cover classes
(\textgreater{} 80\%) for 2001-2022}

@@

For the downloading, processing, and analysis of the spatio-temporal
data, we used the open source software for statistical computing and
graphics, \texttt{R} \citep{R2023}. For downloading ERA5L, we used the
\texttt{\{ecmwfr\}} package \citep{Hufkens2019}. For processing raster
data, we used \texttt{\{terra\}} \citep{Hijmans2023} and
\texttt{\{stars\}} \citep{Pebesma2023}. For managing vectorial data, we
used \texttt{\{sf\}} \citep{Pebesma2018}. For the calculation of AED, we
used \texttt{\{SPEI\}} \citep{Bergueria2023}. For mapping, we use
\{tmap\} \citep{Tennekes2018}. For data analysis, the suite
\{tidyverse\} \citep{Wickham2019} was used.

@@

To derive a proxy for vegetation productivity, we used the product
MOD13A3 collection 6.1 from MODIS \citep{Didan2015}. It provides
vegetation indices (NDVI and EVI) at 1km of spatial resolution and
monthly frequency. The MOD13A3.061 and MCD12Q1.061 were retrieved from
the online Data Pool, courtesy of the NASA EOSDIS Land Processes
Distributed Active Archive Center (LP DAAC), USGS Earth Resources
Observation and Science (EROS) Center, Sioux Falls, South Dakota,
https://lpdaa.usgs.gov/tools/data-pool/.

@@

To evaluate water demand, we chose the \(EDDI\)
\citep{Hobbins2016, McEvoy2016} index, which uses the \(AED\). We used
the SPI \citep{Mckee1993}, a drought monitoring index that the World
Meteorological Organization (WMO) recommends. We calculated the SPEI,
which used a balance between \(P\) and \(AED\), in this case, an
auxiliary variable \(D = P-AED\) is used. In this study, we used the
\(SSI\) (standardized soil moisture index at 1 m)
\citep{Hao2013, AghaKouchak2014}, which uses soil moisture at 1m depth.
Finally, for the proxy of productivity, \(zcNDVI\), we used the NDVI.
Before using the NDVI, it was smoothed using a locally weighted
polynomial regression, following the procedure described in
\citet{Zambrano2018} and \citet{Zambrano2016}.

@@

Climate, vegetation development, seasonality, and changes in vegetation
type all have an impact on the time series of NDVI. In this study, we
want to examine the variation in vegetation productivity across various
land cover types and how water demand, water supply, and soil moisture
affect it. In order to avoid changes due to a change in the land cover
type that will wrongly impact NDVI, we will develop a persistence mask
for land cover for 2001--2023. Thereby, we reduce an important source of
variation on a regional scale.

Therefore, we generated a raster mask for IGBP MODIS per pixel using
macroclasses that remain constant for at least 80\% of the years
(2001--2022). This enables us to identify regions where the land cover
of the macroclasses has remained constant.
